---
title: "Final"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Packages

```{r}
library(tidyverse)
library(rvest)
```

## Data 

```{r, message=FALSE}
masters <- read_csv("Data/Master.csv")
coaches <- read_csv("Data/Coaches.csv")
awardCoaches <- read_csv("Data/AwardsCoaches.csv")
teams <- read_csv("Data/Teams.csv")
teamsPost <- read_csv("Data/TeamsPost.csv")

wiki <- read_html("https://en.wikipedia.org/wiki/List_of_Stanley_Cup_champions")

SC <- html_nodes(wiki, css = "table")

sc_data <- html_table(SC, header = T, fill = T)[[3]]
                 
```

#Tanush
```{r}
#selecting NHL
teams_F <- teams %>% filter(lgID == "NHL")
teamsPost_F <- teamsPost %>% filter(lgID == "NHL")

# | tmID=="LAK" | tmID=="BOS"
teams_F %>% filter(tmID=="MTL") %>% ggplot(aes(year, W/G)) + geom_point(aes(color=(playoff=="SC"))) + geom_line(alpha =0.6) + labs(y="Wins Proportion") + labs(title = "MTL Record")

teams_F %>% filter(playoff =="SC") %>% ggplot(aes(year, W/G)) + geom_point(aes(color=tmID)) + geom_smooth()+ labs(title = "Stanley Cup Winners", y="Proportion of games won")

```

#Zach
```{r}
#Coach Awards Data set

awardCoaches2 <- awardCoaches %>% 
  filter(award == "Jack Adams") %>% 
  select(-note)

masters_names <- masters %>% 
  select(coachID, firstName, lastName)

awardCoaches_f <- left_join(awardCoaches2, masters_names)

#Postseason Data set

teamsPost2 <- teamsPost %>% 
  group_by(year) %>% 
  filter(max(W))
```


#Pete

```{r}
coaches = coaches %>% filter(lgID == "NHL")
coaches1 = coaches %>% 
  filter(lgID == "NHL") %>% 
  group_by(coachID) %>% 
  summarise(number_of_years_in_league = (1+max(year)-min(year)))

coaches2 = coaches %>% 
  filter(lgID == "NHL") %>% 
  group_by(coachID) %>% 
  summarise(reg_overall_win = sum(w), reg_overall_loss = sum(l), reg_overall_tie = sum(t), reg_num_game = sum(g))

coaches3 = coaches %>% 
  filter(lgID == "NHL") %>% 
  group_by(coachID) %>% 
  summarise(post_overall_win = sum(postw), post_overall_loss = sum(postl), post_overall_tie = sum(postt), post_num_game = sum(postg))

coaches4 = coaches1 %>% 
  full_join(coaches2, by = "coachID") %>%
  full_join(coaches3, by = "coachID")

coaches_f = coaches4 %>% 
  mutate(reg_win_percent = (reg_overall_win/reg_num_game), post_win_percent = post_overall_win/post_num_game) %>% 
  select(coachID:reg_overall_tie, reg_win_percent, post_overall_win:post_overall_tie,post_win_percent)

```

#Question 1  
How does the team performance change after the coach wins the Jack Adams award?  
**answer**  
```{r}
#every coach for every year
test1 = coaches %>% 
  full_join(awardCoaches, by = c("coachID","year","lgID")) %>% 
  filter(lgID == "NHL") %>% 
  full_join(teams, by = c("year","tmID","lgID")) %>% 
  mutate(won_award = !is.na(award)) %>% 
  select(coachID,year,tmID,g,w,l,t,award,won_award) %>% 
  filter(is.na(coachID)==FALSE, year >= 1995, year <=2000) %>%                   #doesn't look like jack adams started until 1973
  arrange(year)
test1

#all the coaches that have won an award
test2 <- test1 %>% 
  filter(won_award == TRUE) %>% 
  arrange(year)
test2
 
# test3 = test1 %>% 
#   filter(coachID == test2$coachID) %>% 
#   mutate()

printPlot <- function(tibble) {
  print(ggplot(tibble, aes(year, ((2*W+T)/G))) + geom_line())
  }

for(x in 1:nrow(test2)){
  teams_F %>%  filter(tmID==test2$tmID[x] & year >= test2$year[x] & year <= (test2$year[x]+5)) %>% printPlot(.)
}
# %>% mutate(PtsPtt = (2(W)+T)/G)


# x <- 1
# teams_F %>%  filter(tmID==test2$tmID[x] & year >= test2$year[x] & year <= (test2$year[x]+5)) %>% ggplot(aes(year, ((2*W+T)/G))) + geom_point()
```

 
#Question 2  
How does the team performance change after winning a Stanley Cup?  
**answer**  
In the graph below, the blue line represents the percentage of total possible points per game that the team that won the Stanley Cup the previous year earns. The red line represents the percentage of total possible  points per game for the rest of the league. Therefore, since the average points per game for all but three years since 1987 is higher for the reigning league champions than the rest of the NHL, team performance tends to remain high for teams that have just won a Stanley Cup the year before.  
```{r}
#stantley cup champion, year + 1 to be able to join with another dataset
#to see how they do the next year

#filtering for SC champs in modern playoff structure (needs 16 wins to get Stanley cup)
scChamps <- teamsPost %>% 
  filter(year < 2011, W == 16) %>% 
  select(year, tmID) %>% 
  mutate(year = year + 1)

#filtering reg season data
regSeason <- teams %>%
  select(year, tmID, Pts, G) %>% 
  filter(year > 1987) %>% 
  group_by(year) %>% 
  mutate(meanPtsPct = mean(Pts/(2*G)), PtsPct = Pts/(2*G))

data_f <- left_join(scChamps, regSeason, by = c("year","tmID"))

#MAYBE CHANGE THIS TO PTS %????
ggplot(data_f) +
  geom_line(aes(year, PtsPct), col = "blue") +
  geom_point(aes(year, PtsPct), col = "blue") +
  geom_line(aes(year, meanPtsPct), col = "red") +
  geom_point(aes(year,meanPtsPct), col = "red")
```

```{r}
champs96_00 <- teams %>% 
  filter(year >= 1996, year <= 2000, playoff == "SC") %>% 
  select(year, tmID)

teams %>% 
  filter(year >= 1990) %>% 
  right_join(champs96_00, by = "tmID") %>% 
  select(year.x, tmID, G, Pts, year.y) %>% 
  rename(year = year.x, SCyear = year.y) %>% 
  mutate(PtsPct = Pts / (2*G)) %>% 
  ggplot(aes(year, PtsPct, col = tmID)) +
  geom_line() +
  geom_point(aes(SCyear, PtsPct), col = "black") +
  facet_wrap(~tmID)
```


#Question 3  
What coach coached the same team for the longest number of years?  
Consecutively?  
**answer**  
```{r}

```

#Question 4  
What coach won the most stanley cups?
**answer**  
```{r}
question4 = coaches %>% 
  full_join(awardCoaches, by = c("coachID","year","lgID")) %>% 
  filter(lgID == "NHL") %>% 
  full_join(teams, by = c("year","tmID","lgID")) %>% 
  filter(playoff == "SC" ) %>% 
  select(coachID,year,playoff,tmID) %>% 
  group_by(coachID) %>% 
  tally() %>% 
  filter(is.na(coachID)==FALSE) %>% 
  arrange(desc(n))
question4_f = pete %>% 
  left_join(masters, by = c("coachID")) %>% 
  select(coachID,n,firstName,lastName)
head(question4_f,3)
```

#Question 5  
What coach had the best regular season? 
**answer**  

#Question 6  
Is there any correlation between regular season and post-season performance?  
**answer**  

**answer**  
It does not appear that there is any correlation between regular seson and post-season performance. The correlation coefficient is 0.178 which implies that the relationship is VERY weak.  
```{r}
question6a = coaches %>% 
  mutate(reg_win_percent = (w/g), post_win_percent = postw/postg) %>% 
  filter(reg_win_percent > .5)

ggplot(question6a, aes(reg_win_percent, post_win_percent))+
  geom_point()+
  geom_smooth(method = "lm", se = FALSE )

question6b = coaches %>% 
  mutate(reg_win_percent = (w/g), post_win_percent = postw/postg) %>% 
  filter(reg_win_percent > .5, is.na(reg_win_percent) == FALSE, is.na(post_win_percent) == FALSE)
cor(question6b$post_win_percent,question6b$reg_win_percent)
```
#Question 7  
Coaching tactics that consistently result in a good record?  High PKC / GA (Defensively oriented) or High PPG / GF (Offensively oridented)?  
**answer**  
It appears that there are much more defensively oriented teams than offensively oriented team. This is likely to be due to fact that defensively oriented teams throughout history have had higher mean and median win percentages than offensively oriented team.  
```{r}
question7 <- teams_F %>% 
  mutate(winpercentage = W/G, defense_tendancy = PKC/GA, offense_tendancy = PPG/GF) %>%
  mutate(orientation = ifelse(defense_tendancy > 1.2 | offense_tendancy < .228, "Defense","Offense")) %>% 
  filter(is.na(orientation)==FALSE)

question7i = question7 %>% filter(orientation == "Defense") 
mean(question7i$winpercentage)
median(question7i$winpercentage)
sd(question7i$winpercentage)
question7ii = question7 %>% filter(orientation == "Offense") 
mean(question7ii$winpercentage)
median(question7ii$winpercentage)
sd(question7ii$winpercentage)


ggplot(question7, aes(x = winpercentage))+
  geom_histogram(aes(fill = orientation))
```

